Bow shocks, bow waves, and dust waves. II. Beyond the rip point

### 2019-03-08 papers/Papers1234/app-dust-equations.tex
Major surgery on Paper II

Following \citet{Draine:1979a}, the drag force on a grain that is
moving at relative velocity \(\bm{w} = \bm{v}\grain - \bm{v}\gas\)
through a partially ionized gas can be written as a sum over each
collider species, \(k\), with mass \(m_k\), abundance relative to
hydrogen \(\alpha_k\) and charge \(z_k\). If the relative speed,
\(w = \abs{\bm{w}}\), is normalized to the thermal speed of each
species:
\begin{equation}
  \label{eq:s-velocity}
  s_k = \left( m_k w^2 / 2 k T  \right)^{1/2} \ ,
\end{equation}
then the magnitude of the force is 
\begin{equation}
  \label{eq:ds79}
  f\drag = f_* \sum_k \alpha_k \left[ G_0(s_k) + z_k^2 \phi^2 \ln(\Lambda/z_k) G_2(s_k) \right],
\end{equation}
where \(f_*\) is a characteristic thermal force on the grain (see
eq.~[\ref{eq:fstar}]). The dimensionless functions of normalized speed
\(G_0(s)\) and \(G_2(s)\) are given by
\begin{align}
  \label{eq:G0}
  G_0(s) & = \left( s^2 + 1 - \frac{1}{4 s^2} \right) \erf(s)
  +  \left( s + \frac{1}{2 s} \right) \frac{e^{-s^2}}{\sqrt{\pi}}\\
  \label{eq:G2}
  G_2(s) & = \frac{\erf(s)}{s^2} - \frac{2 e^{-s^2}}{s \sqrt{\pi}} \ .
\end{align}
The \(G_0\) term is due to inelastic solid-body collisions in the
Epstein limit, and is derived in \S~4 of \citet{Baines:1965a}.  The
\(G_2\) term is due to electrostatic Coulomb interactions, with
\(\phi\) being the grain potential in thermal units
(eq.~[\ref{eq:phi-potential}]) and \(\Lambda\) the plasma parameter.  It was
first derived in the different context of dynamical friction in
stellar systems by \citet{Chandrasekhar:1941a}.
Figures~\ref{fig:drag-components}--\ref{fig:gas-grain-drag-photoionized}
show example applications to gas--grain drag in a photoionized region.

The grain trajectories presented in \S~\ref{sec:grain-traj-along} and
\ref{sec:grain-traj-with} are calculated by numerically solving the
grain equation of motion:
\begin{equation}
  \label{eq:grain-equation-motion}
  m\grain \frac{d^2 \bm{r}}{d t^2} = \bm{f} \ .
\end{equation}
The total force \(\bm{f}\) is the sum of radiation, drag, and Lorentz
terms:
\begin{equation}
  \label{eq:total-force}
  \bm{f} = \frac{\sigma\grain \Qp L}{4\pi R^2 c} \hat{\bm{r}}
  - f\drag \hat{\bm{w}}
  + \frac{z\grain e}{c} \bm{w} \times \bm{B} \ ,
\end{equation}
with \(f\drag\) given by equation~\eqref{eq:ds79} and where
\(\hat{\bm{r}}\) is the unit vector in the radial direction and
\(\hat{\bm{w}} = \bm{w} / w\) is the unit vector along the direction
of gas--grain relative motion.  In the strong magnetic coupling limit
(see \S~\ref{sec:grain-traj-with} and \ref{sec:tight-magn-coupl}), the
Lorentz term is not included explicitly, but instead the equation of
motion is solved for the guiding center by replacing \(\bm{f}\) by its
projection along the magnetic field: 
\begin{equation}
  \label{eq:projected-force}
  \widetilde{\bm{f}} = (\bm{f} \cdot \hat{\bm{b}})  \, \hat{\bm{b}} \ ,
\end{equation}
where \(\hat{\bm{b}} = \bm{B} / B\).

If distances are measured in units of the radiative turnaround radius,
\(R\starstar\) (eq.~[\ref{eq:dust-r0}]), and times in units of
\(R\starstar / v_\infty\), then the grain acceleration
\(\bm{a}\grain = \bm{f} / m\grain\) in the non-magnetic case can be
written in non-dimensional form as
\begin{equation}
  \label{eq:grain-acceleration}
  \frac{\bm{a}\grain}{a\starstar}
  = \frac{R\starstar^2}{2 R^{2}} \hat{\bm{r}}
  - C\drag \frac{f\drag}{f_*} \hat{\bm{w}} \, 
\end{equation}
where \(a\starstar = v_\infty^2 / R\starstar\) is a characteristic
acceleration scale and the dimensionless drag constant is
\begin{equation}
  \label{eq:drag-constant}
  C\drag = \frac{4}{\Qp} \left(\frac{\sound \tau_* \kappa\grain}{v_\infty \kappa}\right)^2 \ .
\end{equation}

>>>>

Following \citet{Draine:1979a}, the drag force on a grain that is
moving at relative velocity \(\bm{w} = \bm{v}\grain - \bm{v}\gas\)
through a partially ionized gas can be written as a sum over each
collider species, \(k\), with mass \(m_k\), abundance relative to
hydrogen \(\alpha_k\) and charge \(z_k\). If the relative speed,
\(w = \abs{\bm{w}}\), is normalized to the thermal speed of each
species:
\begin{equation}
  \label{eq:s-velocity}
  s_k = \left( m_k w^2 / 2 k T  \right)^{1/2} \ ,
\end{equation}
then the magnitude of the force is 
\begin{equation}
  \label{eq:ds79}
  f\drag = f_* \sum_k \alpha_k \left[ G_0(s_k) + z_k^2 \phi^2 \ln(\Lambda/z_k) G_2(s_k) \right],
\end{equation}
where \(f_*\) is a characteristic thermal force on the grain (see
eq.~[\ref{eq:fstar}]). The dimensionless functions of normalized speed
\(G_0(s)\) and \(G_2(s)\) are given by
\begin{align}
  \label{eq:G0}
  G_0(s) & = \left( s^2 + 1 - \frac{1}{4 s^2} \right) \erf(s)
  +  \left( s + \frac{1}{2 s} \right) \frac{e^{-s^2}}{\sqrt{\pi}}\\
  \label{eq:G2}
  G_2(s) & = \frac{\erf(s)}{s^2} - \frac{2 e^{-s^2}}{s \sqrt{\pi}} \ .
\end{align}
The \(G_0\) term is due to inelastic solid-body collisions in the
Epstein limit, and is derived in \S~4 of \citet{Baines:1965a}.  The
\(G_2\) term is due to electrostatic Coulomb interactions, with
\(\phi\) being the grain potential in thermal units
(eq.~[\ref{eq:phi-potential}]) and \(\Lambda\) the plasma parameter.  It was
first derived in the different context of dynamical friction in
stellar systems by \citet{Chandrasekhar:1941a}.

Figures~\ref{fig:drag-components}--\ref{fig:gas-grain-drag-photoionized}
show example applications of these equations to gas--grain drag in a
photoionized region.  The included collider species are protons,
electrons, helium ions, and metal ions.  Helium is assumed to be
singly ionized, leading to only a small contribution to the drag
force.  For much hotter stars, such as the central stars of planetary
nebulae, helium may be doubly ionized, which leads to a fourfold
increase in its Coulomb drag contribution, which is significant for
\(w < \SI{5}{km.s^{-1}}\).  All metals are lumped together as a single
species, assuming standard \hii{} region gas-phase abundances.  They
are dominated by C and O, with minor contributions from N and Ne.  The
total abundance is \num{8.5e-4} and the effective atomic weight is
\num{15.3}.  All are assumed to be doubly ionized.  Their largest
relative contribution to the drag force is for
\(w < \SI{2}{km.s^{-1}}\), but is less than 1\% even there.

The grain trajectories presented in \S~\ref{sec:grain-traj-along} and
\ref{sec:grain-traj-with} are calculated by numerically solving the
grain equation of motion:
\begin{equation}
  \label{eq:grain-equation-motion}
  m\grain \frac{d^2 \bm{r}}{d t^2} = \bm{f} \ .
\end{equation}
The total force \(\bm{f}\) is the sum of radiation, drag, and Lorentz
terms:
\begin{equation}
  \label{eq:total-force}
  \bm{f} = \frac{\sigma\grain \Qpbar L}{4\pi R^2 c} \hat{\bm{r}}
  - f\drag \hat{\bm{w}}
  + \frac{z\grain e}{c} \bm{w} \times \bm{B} \ ,
\end{equation}
with \(f\drag\) given by equation~\eqref{eq:ds79} and where
\(\hat{\bm{r}}\) is the unit vector in the radial direction and
\(\hat{\bm{w}} = \bm{w} / w\) is the unit vector along the direction
of gas--grain relative motion.  In the strong magnetic coupling limit
(see \S~\ref{sec:grain-traj-with} and \ref{sec:tight-magn-coupl}), the
Lorentz term is not included explicitly, but instead the equation of
motion is solved for the guiding center by replacing \(\bm{f}\) by its
projection along the magnetic field: 
\begin{equation}
  \label{eq:projected-force}
  \widetilde{\bm{f}} = (\bm{f} \cdot \hat{\bm{b}})  \, \hat{\bm{b}} \ ,
\end{equation}
where \(\hat{\bm{b}} = \bm{B} / B\).

If distances are measured in units of the radiative turnaround radius,
\(R\starstar\) (eq.~[\ref{eq:dust-r0}]), and times in units of
\(R\starstar / v_\infty\), then the grain acceleration
\(\bm{a}\grain = \bm{f} / m\grain\) in the non-magnetic case can be
written in non-dimensional form as
\begin{equation}
  \label{eq:grain-acceleration}
  \frac{\bm{a}\grain}{a\starstar}
  = \frac{R\starstar^2}{2 R^{2}} \hat{\bm{r}}
  - C\drag \frac{f\drag}{f_*} \hat{\bm{w}} \, 
\end{equation}
where \(a\starstar = v_\infty^2 / R\starstar\) is a characteristic
acceleration scale and the dimensionless drag constant is
\begin{equation}
  \label{eq:drag-constant}
  C\drag = \frac{4}{\Qpbar} \left(\frac{\sound \tau_* \kappa\grain}{v_\infty \kappa}\right)^2 \ .
\end{equation}

### 2019-03-08 papers/Papers1234/app-dust-wave.tex
Major surgery on Paper II

We will show later in \S~\ref{sec:magn-effects-grain} that for
sufficiently small grains the Larmor magnetic gyration radius, \(r\B\)
is typically small compared with other length scales of interest, and
so the grains are effectively tied to the magnetic field lines.  In
this approximation, we can calculate the grain dynamics in the
drag-free limit using \(f\rad\) as the sole force as above, but this
time allowing acceleration only along the field lines.  Assuming a
uniform magnetic field, the only extra parameter needed is
\(\theta\B\), the angle between the field direction and the direction of
the dust stream (assumed to be plane parallel).  The radiation force
will also produce an out-of-plane drift, given by
\begin{equation}
  \label{eq:perpendicular-drift}
  \bm{v}_{\text{drift}} = \frac{c}{e z\grain} \, \frac{\bm{f}\!\rad \times \bm{B}}{B^2}
  \ ,
\end{equation}
but from equations~(\ref{eq:Larmor}, \ref{eq:dust-rad-force},
\ref{eq:dust-r0}) it follows that
\begin{equation}
  \label{eq:vdrift-over-vinfinity}
  \frac{v_{\text{drift}}(R\starstar)}{v_\infty} = \frac{r\B}{2 R\starstar} \ ,
\end{equation}
so it is valid to ignore this drift in the limit of small \(r\B\).

We now calculate in detail the grain trajectories in this limit for
two cases, with the magnetic field oriented parallel and perpendicular
to the stream\footnote{In both cases, the stream trajectories are
  assumed parallel to one another, as in \S~\ref{sec:dust-parallel}.}
direction, respectively. These are sufficient to give a flavor of the
effects of a magnetic field on the dust wave structure.  Further
models at intermediate angles, and which also include the effects of
gas drag, are presented in \S~\ref{sec:magn-effects-grain}.

>>>>

We will show later in \S~\ref{sec:magn-effects-grain} that for
sufficiently small grains the Larmor magnetic gyration radius, \(r\B\),
is typically small compared with other length scales of interest, and
so the grains are effectively tied to the magnetic field lines.  In
this approximation, we can calculate the grain dynamics in the
drag-free limit using \(f\rad\) as the sole force as above, but this
time allowing acceleration only along the field lines.  Assuming a
uniform magnetic field, the only extra parameter needed is
\(\theta\B\), the angle between the field direction and the direction of
the dust stream (assumed to be plane parallel).

We now calculate in detail the grain trajectories in this limit for
two cases, with the magnetic field oriented parallel and perpendicular
to the stream direction, respectively. In both cases, the stream
trajectories are assumed parallel to one another, as in
\S~\ref{sec:dust-parallel}. These are sufficient to give a flavor of
the effects of a magnetic field on the dust wave structure.  Further
models at intermediate angles, and which also include the effects of
gas drag, are presented in \S~\ref{sec:magn-effects-grain}.

### 2019-03-08 papers/Papers1234/bs-bw-dw-02.tex
Major surgery on Paper II

>>>>

\section{Introduction}
\label{sec:rip-introduction}

In \S~\ref{sec:gas-free-bow} we calculate analytical
models of the shape of dust waves in the drag-free limit.
% %
% In \S\ref{sec:cloudy-models-dust} we discuss the relevant dust physics
% and the Cloudy models that we employ.
%
In \S~\ref{sec:imperf-coupl-betw} we calculate in detail the grain
charging and dynamics, subject to radiation and drag forces, in order
to determine the \textit{rip point}, which is where gas--grain coupling
catastrophically breaks down.  This is used to determine existence
conditions for the presence of a decoupled dust wave.
%
In \S~\ref{sec:magn-effects-grain} we consider the coupling between
the grains and the plasma's magnetic field, and how this effects the
existence and structure of dust waves.
%
In Appendix~\ref{sec:equat-moti-grains} we provide additional
technical information on the numerical calculation of grain
trajectories.

\input{app-dust-wave}
